Associations of severe COVID-19 with drug prescribing in the REACT-SCOT case-control study

### 2020-05-23 casecontrol.Rmd
fixed coding of severity groups with icu and adm28

---
title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population '
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart J McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Thomas M Caparrotta 
    affiliation: IGMM
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: Public Health Scotland COVID-19 Health Protection Study Group
    affiliation: HPS
address: 
  - code: Usher
    address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care

@@

## Information governance
This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147. Datasets were de-identified before analysis.

@@

## Association with demographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  In comparison with those classified as White, South Asians were at lower risk of severe disease.  <!---Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.  The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.--->

## Associations with listed conditions
Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.  The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[16, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[16, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[16, 8])` of the cases and `r gsub("\\\\", "", table.agegr[16, 7])` of the controls had at least one listed condition.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 age bands combined.  All listed conditions were associated with severe disease in each age band.  In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][6, 3]` for Type 1 diabetes and `r tables.agegr[[1]][7, 3]` for Type 2 diabetes.  For all age groups combined, the rate ratio was `r tables.agegr[[1]][8, 3]` for ischemic heart disease and `r tables.agegr[[1]][9, 3]` for the broad category "other heart disease".  The multivariate analyses shown in Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that in each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.

## Systematic analysis of diagnoses associated with severe disease
The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (Mental) and 10 (musculoskeletal). Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters, filtered to show only subchapters for whcih the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.  This was restricted to those without any listed conditions. This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.  We note that past hospital diagnoses of infections, pneumonia and acute respiratory diseases are strongly associated with severe COVID-19. Associations with cardiovascular disease are not limited to heart disease but are also present for stroke and other circulatory disorders that are not designated as risk conditions.

@@

In contrast to reports from England [@the_opensafely_collaborative_opensafely_2020] we found that people of South Asian ethnicity are at lower risk than Whites.  This suggests that socioeconomic rather than genetic factors underlie at least some of the associations of risk with ethnicity reported in England.  In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].

@@

Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.  However we found that almost half of cases under 40 years had at least one of the listed conditions.  Among the remainder who did not have a listed condition, the proportions who had had at least one hospital admission in the past five years and had encashed at least one prescription in the past year were much higher among cases than in controls.  Thus even among younger persons the vast majority developing  severe disease have  some pre-existing health condition.

@@

# Acknowledgements
We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland and the Public Health Scotland Terminology Services.

@@

# Conflicts of interest
The authors declare no conflicts of interest.

# Supplementary material
The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.

@@

plotroc(full.densities) +
    scale_color_manual(values=c("orange", "purple")) + 
    coord_fixed(ratio=1) +
    theme(axis.title=element_text(size=18)) +
    theme(legend.title=element_blank()) + 
    theme(legend.position=c(0.8, 0.2)) + 
    scale_x_reverse(limits=c(1, 0), expand=c(0, 0)) +
    scale_y_continuous(limits=c(0, 1), expand=c(0, 0)) +
    xlab("Specificity") +
    ylab("Sensitivity (reverse scale)")

@@

knitr::kable(table.demog.aug[, 1:4],
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="1.8cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    group_rows("Ethnicity", 1, 3) %>%
    group_rows("SIMD quintile", 4, 8) %>%
    kable_styling(latex_options="hold_position")

@@

@@

# Appendix
## Supplementary tables

```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}

options(knitr.kable.NA = '')
if(FALSE) {
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             digits=c(0, 0, 0, 200), 
             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity classified in hospital records") %>%
    column_spec(column=1, width="3cm") %>%
    kable_styling(latex_options="hold_position") 
}
```

>>>>

---
title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors for COVID-19 in Scotland (REACT-SCOT): a population-based case-control study'
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart J McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Thomas M Caparrotta 
    affiliation: IGMM
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: Public Health Scotland COVID-19 Health Protection Study Group
    affiliation: HPS
address: 
  - code: Usher
    address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care

@@

@@

## Risk factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  Associations with ethnicity are shown for the full dataset based on name classification.  and separately for the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record.  Over all age groups, `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.  The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[16, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[16, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[16, 8])` of the cases and `r gsub("\\\\", "", table.agegr[16, 7])` of the controls had at least one listed condition.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 year age bands combined.  All listed conditions were associated with severe disease in each age band.  In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][6, 3]` for Type 1 diabetes and `r tables.agegr[[1]][7, 3]` for Type 2 diabetes.  For all age groups combined, the rate ratio was `r tables.agegr[[1]][8, 3]` for ischemic heart disease and `r tables.agegr[[1]][9, 3]` for the broad category "other heart disease".  The multivariate analyses shown in Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that in each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.  Among those aged under 40 years, `r table.scripordiag[2, 2]` of the cases and `r table.scripordiag[2, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year. Among those aged under 60 years, `r table.scripordiag.fatal[2, 2]` of the fatal cases and `r table.scripordiag.fatal[2, 1]` of their matched controls had either a hospital admission in the last five years or a dispensed prescription in the last year.

## Systematic analysis of diagnoses associated with severe disease
The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (mental) and 10 (musculoskeletal). Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters, filtered to show only subchapters for which the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.  This was restricted to those without any listed conditions. This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.  Past hospital diagnoses of infections, pneumonia and acute respiratory diseases were strongly associated with severe COVID-19.  Cardiovascular diagnoses associated with COVID-19 were not limited to heart disease but included also stroke and other circulatory disorders that are not designated as risk conditions.

@@

In contrast to reports from England [@the_opensafely_collaborative_opensafely_2020] we found that people of South Asian ethnicity are at lower risk than Whites.  This suggests that socioeconomic rather than genetic factors underlie at least some of the associations of risk with ethnicity reported in England.  <!---In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020]. Comment - not relevant to South Asians --->

@@

Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.  However we found that almost half of cases under 40 years had at least one of the listed conditions.  Among the remainder who did not have a listed condition, the proportions with at least one hospital admission in the past five years and at least one dispensed prescription in the past year were much higher among cases than in controls.  Thus even among younger persons the vast majority developing severe disease have some pre-existing health condition.

@@

# Declarations
## Information governance
This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147. Datasets were de-identified before analysis.

## Conflicts of interest
The authors declare no conflicts of interest.

## Acknowledgements
We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland and the Public Health Scotland Terminology Services.

@@

## Supplementary material
The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.

@@

plotroc(full.densities) +
    scale_color_manual(values=c("orange", "purple"), labels=c("Crude", "Adjusted")) + 
    coord_fixed(ratio=1) +
    theme(axis.title=element_text(size=18)) +
    theme(legend.title=element_blank()) + 
    theme(legend.position=c(0.8, 0.2)) + 
    scale_x_reverse(limits=c(1, 0), expand=c(0, 0)) +
    scale_y_continuous(limits=c(0, 1), expand=c(0, 0)) +
    xlab("Specificity") +
    ylab("Sensitivity (reverse scale)")

@@

table.demog.append <- table.ethnicsmr.aug
colnames(table.demog.append) <- colnames(table.demog.aug)[1:4]
table.demog.append[, 4] <- as.character(table.demog.append[, 4])
table.demog.aug[, 4] <- as.character(table.demog.aug[, 4])

table.demog.append <- data.frame(varname=c(rownames(table.demog.aug),
                                           rownames(table.ethnicsmr.aug)),
                                 rbind(table.demog.aug[, 1:4], table.demog.append))

knitr::kable(table.demog.append[, 1:5],
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             row.names=FALSE,
             col.names=c("", clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align="lcccc", 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="1.8cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    group_rows("Ethnicity based on name classification", 1, 3) %>%
    group_rows("SIMD quintile", 4, 8) %>%
    group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13,
               hline_before=TRUE, latex_gap_space="10pt") %>% 
    kable_styling(latex_options="hold_position")

@@

```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
if(FALSE) {
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             digits=c(0, 0, 0, 200), 
             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity classified in hospital records") %>%
    column_spec(column=1, width="3cm") %>%
    kable_styling(latex_options="hold_position") 
}
```
\newpage

@@

# Supplementary tables

### 2020-05-23 pharmaco.Rmd
fixed coding of severity groups with icu and adm28

---
title: 'Associations of severe COVID-19 with drug prescribing in the REACT-SCOT case-control study'
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: Public Health Scotland COVID-19 Health Protection Study Group
    affiliation: HPS
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
csl: ../plos.csl
urlcolor: red

@@

## Case ascertainment and sampling of controls 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier. All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.

@@

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge. Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.  British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  For this analysis prescription codes from chapters 14 and above, comprising vaccines, anaesthesia, other preparations, dressings and appliances, were grouped as "Other". Most of these codes are for dressings and appliances rather than drugs.

@@

## Information governance
This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.

@@

This led us to examine further the association of proton pump inhibitors with severe disease, without restricting the analysis to those not resident in care homes and without a diagnosis of a listed condition.  Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".  The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates.

@@

@@

```{r cats3, echo=FALSE, warning=FALSE, message=FALSE}

@@

>>>>

---
title: 'Associations of severe COVID-19 with drug prescribing in the REACT-SCOT case-control study'
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Sharon Kennedy
    affiliation: HPS
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: Public Health Scotland COVID-19 Health Protection Study Group
    affiliation: HPS
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
csl: ../plos.csl
urlcolor: red

@@

## Case ascertainment and sampling of controls 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier. All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained for this study.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.

@@

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge. Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.  British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding those dispensed less than 25 days before testing positive for SARS-CoV-2.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  For this analysis prescription codes from chapters 14 and above, comprising vaccines, anaesthesia, other preparations, dressings and appliances, were grouped as "Other". Most of these codes are for dressings and appliances rather than drugs.  Defined daily doses (DDDs) were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.  Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD)  the time in

@@

@@

This led us to examine further the association of proton pump inhibitors with severe disease, without restricting the analysis to those not resident in care homes and without a diagnosis of a listed condition.  Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".  The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates.  Table \ref{tab:dose.protonpump} shows how the effect size varies with the dose of proton pump inibitor, calculated as average defined daily doses dispensed per day since 1 June 2019.  There is a strong dose-response relationship in the two younger age groups, but not in those aged over 75 years.

@@

# Declarations
## Information governance
This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.

@@

```{r cats3, echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')

@@

```{r dose.protonpump, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.dose.protonpump,
             escape=FALSE,
             label="dose.protonpump", 
             booktabs=TRUE,
             row.names=FALSE,
             col.names=c("Average DDDs/day", "Controls", "Cases",
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe disease with average dose of proton pump inhibitor (daily defined doses/day) dispensed since 1 June 2019, by age group") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    group_rows("Ages 0-59 years", 1, 5) %>%
    group_rows("Ages 60-74 years", 6, 10) %>%
    group_rows("Ages 75+ years", 11, 15) %>%
    landscape()
